---
title: "Statistical Modelling Report"
author: "Nisha Daniel"
output:
  pdf_document: default
  word_document: default
---

Word Count:   lines words bytes chars nonws
               177   849  5698  5692  4557
          

# **Quit Smoking! Would You Like to have Vegetables and Fish or Red
Meat?**

Lifestyle behaviours can win or lose life itself. The following
statistical analysis supports a research on Colorectal Cancer(CRC)
prevalence to build a statistical model that identify healthy lifestyle
behaviours and to suggest any modification to hazardous habits, based on
the real-world Case Control Colorectal Cancer dataset collected in
Scotland.

## Pre-preparations

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#Loading Libraries
library(tinytex)
library(tidyverse)
library(Hmisc)
library(gtsummary)
library(cowplot)


#Reading in the data
crc_original <- read_csv("Case Control CRC(1).csv")
view(crc_original)
#check for missing data
colSums(is.na(crc_original))

```

## DATA WRANGLING

After reading in the data and checking for any missing data, all
categorical variables are recoded into factors. Age variable is treated
as an integer.

```{r CRC}
categories = c("Group", "Sex", "FamHist", "DepCat", "Smoking", "RedMeat", "Veg", "Fish")
crc <-
crc_original %>%
mutate(Group = factor(Group, levels = c("Control", "CRC"), labels = c("Control", "CRC")),
FamHist = factor(FamHist, levels = c(1, 2, 3), labels = c("No", "Immediate", "Distant")),
Sex = factor(Sex, levels = c(1, 2), labels = c("Male", "Female")),
DepCatC = factor(DepCat, levels = c(1:7), labels = c(1:7)),
Smoking = factor(Smoking, levels = c(1 , 2 , 3), labels = c("Non", "Current", "Ex")),
RedMeat = factor(RedMeat, levels = c(1, 2), labels = c("High", "Low")),
Veg = factor(Veg, levels = c(1, 2), labels = c("High", "Low")),
Fish = factor(Fish, levels = c(1, 2), labels = c("High", "Low")),
Age = as.integer(Age))

summary(crc)

```

# DATA EXPLORATION USING SIMPLE DESCRIPTIVE METHODS(numericals and graphical)

## Data Visualization of Distributon of Age

Let us explore the variable, Age, to see any relationship with the
outcome variable, CRC.

```{r AgePlotGrid}

plot1 <- crc %>%
ggplot()+
geom_histogram(aes(Age), bins = 15, colour = "grey", fill = "green") +
geom_vline(aes(xintercept = mean(Age)), colour = 'yellow', size = 1, show.legend = TRUE) +
geom_vline(aes(xintercept = median(Age)), colour = 'red', size = 1, show.legend = TRUE) +
theme_bw() +
labs(title = "Age Distribution", )

plot2 <- crc %>%
ggplot() +
geom_boxplot(aes(Group, y = Age, colour = Group)) +
theme_bw() +
labs(title = "Age Distribution by Group")

plot_grid(plot1, plot2)
```

The mean and median of Age is above 50 years, citing the risky age
group.A two sample T-test is further done in hypothesis testing of Age,
to see if there's a significant difference between the means of Age, as
it is a continuous variable.

```{r test}
t.test(Age ~ Group, data = crc)

```

The mean age in the CRC group (63.194) is significantly higher than the
control group (50.218). Therefore, **it is likely that the Age variable
is required in the model.**

A categorical variable called AgeQ is created for further analysis
checking the linear relationship.

```{r Age1}
crc <- crc %>%
mutate(AgeQ =cut2(Age,g=4))
table(crc$AgeQ)

summary(crc$Age)

```

Although parameters are not increasing in equal increments, there is
still a general increasing trend visible. For simplicity within the
model, age will be kept as a continuous variable.

### Smoking as a variable To get the proportion of different types of smokers

```{r CheckSmokers}
prop.table(table(crc$Smoking))
```

## Age:Smoke

### Visualizing Interation of Age:Smoking status of the participants

The interaction between Age and Smoking habit is studied and plotted.

```{r SmokePlot}

ggplot(crc, aes(x = Smoking, y = Age)) +  # Map Smoking to x-axis, Age to y-axis
  geom_boxplot(aes(color = Smoking)) +  # Color by Smoking category
  labs(title = "Smoking vs. Age", x = "Smoking Status", y = "Age") +
  theme_classic()
```

A finding from the boxplot chart is that, with high age, there is
increased smoking habit, as observed across the three categories.
Further performing Chi Square test, as Smoking is grouped as a
categorical variable in the dataset.

```{r Chitest Smoke}
#create contingent  table
contingentb_smoke_group <- as.data.frame.matrix(table(crc$Smoking, crc$Age))
#perform chi square test
chisq.test(contingentb_smoke_group)
```

With a low p-value, there is a statistically significant difference in
the rates of colorectal cancer between different smoking categories.
**This variable, along with the interaction, will feature in the final
model.**

```{r other variables}
#Enlist all variables with p-values
crc %>%
tbl_summary(include = c(Sex, Age, FamHist, DepCat, Smoking, RedMeat, Veg, Fish),
by = Group, percent = "row",
label = list(Age = "Age in years",
FamHist = "Family History",
DepCat = "Deprivation Category",
RedMeat = "Red Meat Consumpiton",
Veg = "Vegetable Consumption",
Fish = "Fish Consumption")) %>%
bold_labels() %>%
add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3))
```

# CHOICE OF METHODOLOGY *Logistic Regression Model*

A logistic regression model is the most suitable as the outcome variable
is binary. The model is created using the step function, starting with
forward selection. The function will then add and remove variables in a
stepwise fashion, until the lowest possible AIC is attained.

# MODEL FITTING

## Deriving the Final Model

```{r Model}
crc_model <- step(glm(Group ~ Age,
family = binomial("logit"),
data = crc),
scope = list(lower = .~1,
upper = .~. + Sex + FamHist + DepCat + Smoking + RedMeat + Veg + Fish,
direction = "forward"))
summary(crc_model)
```

## Analysing Variance

```{r ANOVA}
anova(crc_model, test = "Chisq")

```

Following the results of ANOVA, two functions are used to fine-tune the
model.

### a. Using drop1

```{r drop1}
drop1(crc_model, test = "Chisq")
```

### b. Using cbind

```{r cbind}
cbind(OR = exp(crc_model$coefficients),
exp(confint(crc_model)))
```

## Tabulating the Key parameters

```{r table}

tbl_regression(crc_model, exponentiate = TRUE, message = FALSE,
label = list(Age = "Age Category",
FamHist = "Family History",
RedMeat = "Red Meat Intake",
Veg = "Veg diet",
Fish = "Fish Intake")) %>%
bold_labels()

```

# MODEL INTERPRETATION

The final statistical model is arrived at. The following prognostic
variables contribute significantly to colorectal cancer risk: Age,
Family History, Smoking Status and Red Meat consumption, individually.
There is a positive interaction between the Age and Smoking variables
with the outcome variable, which is also included in the model. There is
low effects for Sex and Deprivation status. Nevertheless, the dietary
habits like following vegan diet, intake of fish or red meat or even
conducting hazardous habits are well within the control of the human
participants. Hence, considering only the variables within the ‘locus of
control’ of the participants, a separate sub-model is co-created for the
**Public** using four variables.

### Model to the BioStatistician / Researcher:

The final model is: Group \~ Age + Smoking + FamHist + RedMeat +
Age:Smoking

### Model for Public Health Awareness for the Public:

The final model is: Group \~ Smoking(**remove habit**) +
RedMeat(**decrease intake**) + Veg(**increase intake**) +
Fish(**increase intake**) + Sex(**women at higher risk**)

## Summary of Recommendations:

1.  Compared to non-smokers, current smokers increase the risk of CRC by
    2.34x, whereas ex-smokers also have an increased risk of 1.38x. \n
2.  With each year increase in age, the risk of CRC increases by 1.13x.
    \n
3.  Combining Age and Smoking is a hig risk indicator for people above
    40 years.
4.  In terms of diet, a low red meat diet lowers one’s risk of
    colorectal cancer, with a odds raio of 0.61x, as against Veg diet
    and fish consumption. \n
5.  Family history is a necessary contributor to the development of CRC
    with an odds ratio range of 0.77x to 9.5x, nonetheless, it is out of
    the control of the patient.

## Bibliography:

1.Petrie, A., & Sabin, C. (2020). Medical statistics at a glance (4th
ed.). Wiley-Blackwell.\n  2.Kirkwood, B.R. and Sterne, J.A.C. (2003)
Essential Medical Statistics. 2nd Edition, Blackwell Science, Oxford \n
3.Janet L. Peacock and Phil J. Peacock () Oxford Handbook of Medical
Statistics. 2nd Edition, Oxford University Press

## Terminiologies Explained:

*Colorectal Cancer(CRC):*Colorectal cancer is a type of cancer that
starts in the colon or the rectum. The colon is the large intestine, and
the rectum is the lower part of the large intestine. /

*Prevalence:*In the context of healthcare and epidemiology, prevalence
refers to the proportion of a population that has a specific disease or
characteristic at a particular point in time.
